---
title: "02 Pattern Analysis"
author: "Fu Wei Hsu"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: right
    toc-title: "On this page"
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
---

# Executive Summary

**Objective:**

To conduct a systematic exploration of formatting patterns and special characters across the three datasets, ensuring a unified cleaning protocol prior to data integration.

# importing 01 cleaned data

```{r}
library(tidyverse)
library(janitor)
library(gt) 

# Load cleaned datasets from 01_data_exploration
D1_clean_base <- readRDS("../Data/cleaned_D1.rds")
D2_clean_base <- readRDS("../Data/cleaned_D2.rds")
D3_clean_base <- readRDS("../Data/cleaned_D3.rds")

cat("Datasets loaded successfully\n")
cat("D1 (Billboard):", nrow(D1_clean_base), "records\n")
cat("D2 (Spotify):", nrow(D2_clean_base), "records\n")
cat("D3 (Music Dataset):", nrow(D3_clean_base), "records\n")
```

------------------------------------------------------------------------

# Part 1: Dataset 1 (Billboard) Pattern Analysis

## 1.1 Artist Column Pattern Analysis

```{r}
# Define patterns to analyze
d1_artist_stats <- tibble(
  Pattern = c(
    "Contains 'Featuring'",
    "Contains 'feat.' or 'ft.'",
    "Contains '&'",
    "Contains ','",
    "Contains '()' or '[]'",
    "Contains '/'",
    "Contains '\\'",
    "Contains 'x' or 'X' (Collab)",
    "Contains '+' (Collab)"
  ),
  Count = c(
    sum(str_detect(D1_clean_base$artist, regex("Featuring", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$artist, regex("feat\\.|ft\\.", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$artist, "&")),
    sum(str_detect(D1_clean_base$artist, ",")),
    sum(str_detect(D1_clean_base$artist, "[\\(\\)\\[\\]]")),
    sum(str_detect(D1_clean_base$artist, "/")),
    sum(str_detect(D1_clean_base$artist, "\\\\")),
    sum(str_detect(D1_clean_base$artist, regex("\\sx\\s", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$artist, regex("\\s\\+\\s", ignore_case = TRUE)))
  )
)

# Display table
d1_artist_stats %>%
  gt() %>%
  tab_header(
    title = "Artist Column Pattern Statistics (D1)"
  ) %>%
  cols_label(
    Pattern = "Pattern Description"
  ) %>%
  fmt_number(
    columns = Count,
    decimals = 0,
    use_seps = TRUE 
  )

```

### Representative Examples

```{r}
# define a function to avoid repeating the filtering logic
show_examples <- function(data, pattern_name, regex_pattern) {
  cat(paste0("\nExample: contain ", pattern_name, "\n"))
  
  data %>%
    filter(str_detect(artist, regex(regex_pattern, ignore_case = TRUE))) %>%
    distinct(artist) %>%
    slice_head(n = 2) %>%
    pull(artist) %>%
    print()
}

show_examples(D1_clean_base, "Featuring", "featuring")
show_examples(D1_clean_base, "feat./ft.", "feat\\.|ft\\.")
show_examples(D1_clean_base, "Ampersand (&)", "&")
show_examples(D1_clean_base, "Comma (,)", ",")
show_examples(D1_clean_base, "Brackets () []", "[\\(\\)\\[\\]]")
show_examples(D1_clean_base, "Forward Slash (/)", "/")
show_examples(D1_clean_base, "Backslash (\\)", "\\\\")
show_examples(D1_clean_base, "x or X (Collab)", "\\sx\\s")
show_examples(D1_clean_base, "Plus (+)", "\\s\\+\\s")

```

### Deep Analysis

##### Analysis 1: Usage Context of '&' Symbol

```{r}
ampersand_data <- D1_clean_base %>% 
  filter(str_detect(artist, "&")) %>% 
  distinct(artist) %>% 
  mutate(
    # Check for collaboration markers
    has_collab_marker = str_detect(artist, 
      regex("featuring|feat\\.|ft\\.|with", ignore_case = TRUE)),
    
    # Count occurrences of '&'
    ampersand_count = str_count(artist, "&"),
    
    # Categorize
    type = case_when(
      has_collab_marker ~ "Has collab marker (Safe to clean)",
      ampersand_count >= 2 ~ "Multiple '&' (Complex collab)",
      TRUE ~ "Simple '&' (Likely duo or group)"
    )
  )

cat("Total artists containing '&':", nrow(ampersand_data), "\n\n")

# Summary
type_summary <- ampersand_data %>% 
  count(type) %>% 
  mutate(Percentage = round(n / nrow(ampersand_data) * 100, 1))

type_summary %>% 
  gt() %>% 
  tab_header(title = "Analysis of '&' Symbol Usage Types") %>%
  fmt_number(columns = n, decimals = 0, use_seps = TRUE) 

```

##### Analysis 1: Representative Examples

```{r}
# Define a helper function to display samples
display_samples <- function(data, type_name, title, strategy) {
  cat("\nCategory:", title, "\n")
  cat("Cleaning Strategy:", strategy, "\n\n")
  
  data %>% 
    filter(type == type_name) %>% 
    slice_head(n = 2) %>% 
    pull(artist) %>% 
    print()
}

# Display samples based on categories
display_samples(ampersand_data, 
                "Has collab marker (Safe to clean)", 
                "Collab marker detected (featuring/with/feat)", 
                "Remove all content following the collab marker.")

display_samples(ampersand_data, 
                "Simple '&' (Likely duo or group)", 
                "Simple '&' (No collab marker)", 
                "Conservative approach; retain the primary artist before '&'.")

display_samples(ampersand_data, 
                "Multiple '&' (Complex collab)", 
                "Multiple '&' (Complex cases)", 
                "Similar to Category 1; remove content after the collab marker.")
```

##### Analysis 1: Cleaning Strategy

```{r}
cat("\n========================================\n")
cat("  Cleaning Strategy for '&' Symbol\n")
cat("========================================\n\n")

cat("   Conservative Cleaning\n")
cat("   - Remove all content after 'featuring', 'feat.', 'ft.', or 'with'.\n")
cat("   - Preserve '&' (do not remove).\n\n")

cat("   Examples:\n")
cat("   'Drake Featuring Future & Young Thug' -> 'drake'\n")
cat("   'The Kid LAROI & Justin Bieber'       -> 'the kid laroi & justin bieber' (Retained)\n")
```

##### Analysis 2: Parentheses Content Analysis

```{r}
# Prepare the data
paren_data <- D1_clean_base %>%
  filter(str_detect(artist, "\\(.*\\)")) %>%
  mutate(
    inside_paren = str_extract(artist, "\\(([^)]+)\\)"),
    inside_paren_clean = str_remove_all(inside_paren, "[\\(\\)]"),
    # Create categories to enable the display_samples function
    type = if_else(str_detect(inside_paren_clean, regex("feat|duet|featuring", ignore_case = TRUE)),
                   "Collab-related (Safe to remove)",
                   "Group Info/Metadata (Preserve)")
  )

cat("Total artists containing parentheses:", nrow(paren_data), "\n\n")

# Summary table of content inside parentheses
paren_data %>%
  count(inside_paren_clean, sort = TRUE) %>%
  slice_head(n = 10) %>%
  gt() %>%
  tab_header(title = "Distribution of Content inside Parentheses") %>%
  fmt_number(columns = n, decimals = 0, use_seps = TRUE)
```

##### Analysis 2: Representative Examples

```{r}
display_samples(paren_data, 
                "Collab-related (Safe to remove)", 
                "Collab-related (featuring/duet)", 
                "Remove parentheses and content.")

display_samples(paren_data, 
                "Group Info/Metadata (Preserve)", 
                "Group Names/Metadata", 
                "Retain full string (preserve for context).")

```

##### Analysis 2: Cleaning Strategy

```{r}
cat("\n========================================\n")
cat("Cleaning Strategy for Parentheses\n")
cat("========================================\n\n")

cat("   Sective Removal\n")
cat("   - Remove parentheses containing 'featuring', 'feat.', or 'duet with'.\n")
cat("   - Preserve other parentheses (e.g., band members, full group names).\n\n")

cat("   Examples:\n")
cat("   'Artist (Featuring Guest)'              -> 'artist' (Removed)\n")
cat("   'Artist (Duet With Guest)'              -> 'artist' (Removed)\n")
cat("   'Silk Sonic (Bruno Mars & Anderson .Paak)' -> 'Silk Sonic (Bruno Mars & Anderson .Paak)' (Preserved)\n")
cat("   'F.L.Y. (Fast Life Yungstaz)'           -> 'F.L.Y. (Fast Life Yungstaz)' (Preserved)\n")

```

#### Analysis 3: Comma Usage Patterns

```{r}

comma_data <- D1_clean_base %>% 
  filter(str_detect(artist, ",")) %>% 
  distinct(artist) %>% 
  mutate(
    # Check for collaboration markers
    has_collab_marker = str_detect(artist, 
      regex("featuring|feat\\.|ft\\.|with", ignore_case = TRUE)),
    
    # Count commas
    comma_count = str_count(artist, ","),
    
    # Categorize
    type = case_when(
      has_collab_marker ~ "Collab marker (Automated processing)",
      comma_count >= 2 ~ "Multiple commas (Complex cases)",
      TRUE ~ "Simple comma (Group/Part of name)"
    )
  )

cat("Total artists containing commas:", nrow(comma_data), "\n\n")

# Summary table
comma_data %>% 
  count(type) %>% 
  mutate(Percentage = round(n / sum(n) * 100, 2)) %>% 
  gt() %>% 
  tab_header(title = "Usage Context Statistics for Commas") %>%
  fmt_number(columns = n, decimals = 0, use_seps = TRUE)
```

##### Analysis 3: Representative Examples

```{r}
# Using the helper function defined in previous steps
display_samples(comma_data, 
                "Collab marker (Automated processing)", 
                "Collab Marker Detected", 
                "Remove collab marker and following content; comma will be handled automatically.")

display_samples(comma_data, 
                "Simple comma (Group/Part of name)", 
                "Simple Comma (No collab marker)", 
                "Preserve fully; likely multiple artists or part of the name.")

display_samples(comma_data, 
                "Multiple commas (Complex cases)", 
                "Multiple Commas (Complex)", 
                "Similar to collab cases; remove following content if collab marker exists.")

```

#### Analysis 3: Cleaning Strategy

```{r}
# Check for ", The" pattern (common in names like "Tyler, The Creator")
name_comma <- D1_clean_base %>% 
  filter(str_detect(artist, ", The")) %>% 
  distinct(artist) %>% 
  slice_head(n = 10)

cat("Artists containing ', The' (likely part of the name):\n")
name_comma %>% 
  pull(artist) %>% 
  print()

cat("These commas are part of the artist's name is not be removed\n")
```

------------------------------------------------------------------------

## 1.2 Song Column Pattern Analysis

```{r}
d1_song_stats <- tibble(
  Pattern = c(
    "Contains Parentheses ()",
    "Contains Brackets []",
    "Contains ' - ' (dash)",
    "Contains 'feat.'",
    "Contains 'Remaster'",
    "Contains 'Live'",
    "Contains 'Remix'",
    "Contains 'Radio Edit'",
    "Contains 'feat' inside ()",
    "Contains '+' (Part of song title)",
    "Contains 'Pt.' or 'Part'",
    "Contains 'Take'"
  ),
  Count = c(
    sum(str_detect(D1_clean_base$song, "\\(.*\\)")),
    sum(str_detect(D1_clean_base$song, "\\[.*\\]")),
    sum(str_detect(D1_clean_base$song, " - ")),
    sum(str_detect(D1_clean_base$song, regex("feat\\.", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$song, regex("Remaster", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$song, regex("Live", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$song, regex("Remix", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$song, regex("Radio Edit", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$song, regex("\\(.*feat", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$song, regex("\\s\\+\\s", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$song, regex("Pt\\.|Part", ignore_case = TRUE))),
    sum(str_detect(D1_clean_base$song, regex("Take", ignore_case = TRUE)))
  )
) %>% 
  mutate(Percentage = round(Count / nrow(D1_clean_base) * 100, 1))

# Display table
d1_song_stats %>% 
  gt() %>% 
  tab_header(title = "D1 Song Column Pattern Statistics") %>% 
  fmt_number(columns = Count, decimals = 0, use_seps = TRUE)
```

### Pattern Examples

```{r}
show_song_examples <- function(data, pattern_regex, label) {
  cat("\n=== Example:", label, "===\n")
  data %>% 
    filter(str_detect(song, regex(pattern_regex, ignore_case = TRUE))) %>% 
    distinct(song) %>% 
    slice_head(n = 3) %>% 
    pull(song) %>% 
    print()
}

# Run the function for each pattern
show_song_examples(D1_clean_base, "\\(.*\\)", "Contains Parentheses ()")
show_song_examples(D1_clean_base, "\\[.*\\]", "Contains Brackets []")
show_song_examples(D1_clean_base, " - ", "Contains ' - ' (dash)")
show_song_examples(D1_clean_base, "feat\\.", "Contains 'feat.'")
show_song_examples(D1_clean_base, "Remaster", "Contains 'Remaster'")
show_song_examples(D1_clean_base, "Live", "Contains 'Live'")
show_song_examples(D1_clean_base, "Remix", "Contains 'Remix'")
show_song_examples(D1_clean_base, "Radio Edit", "Contains 'Radio Edit'")
show_song_examples(D1_clean_base, "\\(.*feat", "Contains 'feat' inside ()")
show_song_examples(D1_clean_base, "\\s\\+\\s", "Contains '+' (Part of title)")
show_song_examples(D1_clean_base, "Pt\\.|Part", "Contains 'Pt.' or 'Part'")
show_song_examples(D1_clean_base, "Take", "Contains 'Take'")

```

### Deep Analysis

#### Analysis 1: Parentheses Content Analysis

```{r}
# Extract and categorize content inside parentheses
paren_data <- D1_clean_base %>%
  filter(str_detect(song, "\\(.*\\)")) %>%
  distinct(song) %>%
  mutate(
    inside_paren = str_extract(song, "\\(([^)]+)\\)"),
    inside_clean = str_remove_all(inside_paren, "[\\(\\)]"),
    
    # Categorization
    type = case_when(
      str_detect(inside_clean, regex("remix", ignore_case = TRUE)) ~ "Remix Version",
      str_detect(inside_clean, regex("^live$|live version|acoustic|unplugged", ignore_case = TRUE)) ~ "Performance Version",
      str_detect(inside_clean, regex("radio edit|album version|single version", ignore_case = TRUE)) ~ "Release Version Tag",
      str_detect(inside_clean, regex("remaster", ignore_case = TRUE)) ~ "Remastered Version",
      str_detect(inside_clean, regex("feat\\.|featuring", ignore_case = TRUE)) ~ "Featuring Info",
      TRUE ~ "Subtitle or Title Part (Keep)"
    )
  )

cat("Total songs containing parentheses:", nrow(paren_data), "\n\n")

# Statistics table
paren_data %>%
  count(type) %>%
  arrange(desc(n)) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1)) %>%
  gt() %>%
  tab_header(title = "Parentheses Content Classification") %>%
  fmt_number(columns = n, decimals = 0, use_seps = TRUE)
```

#### Analysis 1: Representative Examples

```{r}
# Define a cleaner display function
display_samples <- function(data, category_name) {
  cat("\n=== Category:", category_name, "===\n")
  data %>%
    filter(type == category_name) %>%
    slice_head(n = 8) %>%
    pull(song) %>%
    {cat(paste0("  ", seq_along(.), ". '", ., "'\n"), sep = "")}
}

# Run display for each category
display_samples(paren_data, "Remix Version")
display_samples(paren_data, "Performance Version")
display_samples(paren_data, "Release Version Tag")
display_samples(paren_data, "Remastered Version")
display_samples(paren_data, "Featuring Info")
display_samples(paren_data, "Subtitle or Title Part (Keep)")
```

**Top 20 Parentheses Content**

```{r}
paren_data %>%
  count(inside_clean, sort = TRUE) %>%
  slice_head(n = 20) %>%
  gt() %>%
  tab_header(title = "Top 20 Most Frequent Parentheses Content") %>%
  fmt_number(columns = n, decimals = 0, use_seps = TRUE)
```

------------------------------------------------------------------------

#### Analysis 2: Dash Content Analysis

```{r}
# Extract and categorize content after the dash
dash_data <- D1_clean_base %>%
  filter(str_detect(song, " - ")) %>%
  distinct(song) %>%
  mutate(
    # Extract content after the dash
    after_dash = str_extract(song, " - (.*)$"),
    after_dash_clean = str_remove(after_dash, "^ - "),
    
    # Categorization
    type = case_when(
      str_detect(after_dash_clean, regex("remix", ignore_case = TRUE)) ~ "Remix Version",
      str_detect(after_dash_clean, regex("remaster", ignore_case = TRUE)) ~ "Remastered Version",
      str_detect(after_dash_clean, regex("live|acoustic", ignore_case = TRUE)) ~ "Performance Version",
      str_detect(after_dash_clean, regex("radio edit|album version|single", ignore_case = TRUE)) ~ "Release Version Tag",
      TRUE ~ "Subtitle or Title Part (Keep)"
    )
  )

cat("Total songs containing dashes:", nrow(dash_data), "\n\n")

# Statistics table
dash_data %>%
  count(type) %>%
  arrange(desc(n)) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1)) %>%
  gt() %>%
  tab_header(title = "Dash Content Classification") %>%
  fmt_number(columns = n, decimals = 0, use_seps = TRUE)
```

#### Analysis 2: Representative Examples

```{r}
# display_samples(dash_data, "Category Name")

display_samples(dash_data, "Remix Version")
display_samples(dash_data, "Remastered Version")
display_samples(dash_data, "Performance Version")
display_samples(dash_data, "Release Version Tag")
display_samples(dash_data, "Subtitle or Title Part (Keep)")
```

**Top 20 Dash Content**

```{r}
dash_data %>%
  count(after_dash_clean, sort = TRUE) %>%
  slice_head(n = 20) %>%
  gt() %>%
  tab_header(title = "Top 20 Most Frequent Dash Content") %>%
  fmt_number(columns = n, decimals = 0, use_seps = TRUE)
```

------------------------------------------------------------------------

#### Analysis 3: Bracket Content Analysis

```{r}
# Extract content inside brackets
bracket_data <- D1_clean_base %>%
  filter(str_detect(song, "\\[.*\\]")) %>%
  distinct(song) %>%
  mutate(
    inside_bracket = str_extract(song, "\\[([^]]+)\\]"),
    inside_clean = str_remove_all(inside_bracket, "[\\[\\]]")
  )

cat("Total songs containing brackets:", nrow(bracket_data), "\n\n")

# Statistics table
bracket_data %>%
  count(inside_clean, sort = TRUE) %>%
  slice_head(n = 10) %>%
  gt() %>%
  tab_header(title = "Frequency of Content Inside Brackets") %>%
  fmt_number(columns = n, decimals = 0, use_seps = TRUE)
```

#### Analysis 3: Representative Examples

```{r}
bracket_data %>%
  slice_head(n = 10) %>%
  pull(song) %>%
  {cat(paste0("  ", seq_along(.), ". '", ., "'\n"), sep = "")}
```

------------------------------------------------------------------------

#### False Positive Checks

```{r}
# Check 'Live'
## 'Live' as a lyric word
live_in_lyrics <- D1_clean_base %>%
  filter(str_detect(song, regex("\\blive\\b", ignore_case = TRUE))) %>%
  filter(!str_detect(song, regex("\\(.*live.*\\)| - .*live", ignore_case = TRUE))) %>%
  distinct(song) %>%
  slice_head(n = 10)

## 'Live' as part of the song title (Not a Live version)
live_in_lyrics %>%
  pull(song) %>%
  {cat(paste0("  ", seq_along(.), ". '", ., "'\n"), sep = "")}

cat("Using regex('live') would result in false positives!\n")

# Check 'Part'
cat("\n─── Check 2: 'Part' as a lyric word ───\n\n")
party_words <- D1_clean_base %>%
  filter(str_detect(song, regex("party|part", ignore_case = TRUE))) %>%
  filter(!str_detect(song, regex("part\\s+\\d|pt\\.\\s*\\d", ignore_case = TRUE))) %>%
  distinct(song) %>%
  slice_head(n = 10)

## 'Party/Part' as part of the song title
party_words %>%
  pull(song) %>%
  {cat(paste0("  ", seq_along(.), ". '", ., "'\n"), sep = "")}

# Check 'Take'
## 'Take' as a lyric word
take_words <- D1_clean_base %>%
  filter(str_detect(song, regex("\\btake\\b", ignore_case = TRUE))) %>%
  filter(!str_detect(song, regex("take\\s+\\d", ignore_case = TRUE))) %>%
  distinct(song) %>%
  slice_head(n = 10)

## 'Take' as part of the song title
take_words %>%
  pull(song) %>%
  {cat(paste0("  ", seq_along(.), ". '", ., "'\n"), sep = "")}
```

#### Cleaning Strategy

```{r}
cat("\n========================================\n")
cat("D1 Song Column Cleaning Strategy\n")
cat("========================================\n\n")

cat("Extreme Conservative Cleaning - Remove only explicit version tags.\n\n")

cat("Data-Driven Rationale:\n")
cat("  - Parentheses (): 99.40% are part of the song title -> Preserve most.\n")
cat("  - Dash -:        98.96% are part of the song title -> Preserve most.\n")
cat("  - Brackets []:  100% are descriptive metadata    -> Remove all.\n\n")

cat("Cleaning Objectives:\n")
cat("  1. Remove explicit version tags (Remix, Remaster, Live Version, etc.).\n")
cat("  2. Remove descriptive brackets [].\n")
cat("  3. Preserve song subtitles and full titles.\n")
cat("  4. Avoid removing common words in lyrics (Live, Part, Take).\n\n")

cat("Exclusions (I Do NOT clean):\n")
cat("  - Standard parentheses (likely part of the title).\n")
cat("  - Standard dashes (likely subtitles).\n")
cat("  - Words like 'live', 'part', 'take' (usually part of the song lyrics).\n")
```

# Part 2: Dataset 2 (Spotify) Pattern Analysis

## 2.1 Artist Column Pattern Analysis

```{r}
d2_artist_stats <- tibble(
  Pattern = c(
    "Contains 'Featuring'",
    "Contains 'feat.' or 'ft.'",
    "Contains '&'",
    "Contains comma ','",
    "Contains brackets () or []",
    "Contains forward slash /",
    "Contains backslash \\",
    "Contains 'x' or 'X' (Collab marker)",
    "Contains '+' (Collab marker)"
  ),
  Count = c(
    sum(str_detect(D2_clean_base$artist, regex("Featuring", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$artist, regex("feat\\.|ft\\.", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$artist, "&")),
    sum(str_detect(D2_clean_base$artist, ",")),
    sum(str_detect(D2_clean_base$artist, "[\\(\\)\\[\\]]")),
    sum(str_detect(D2_clean_base$artist, "/")),
    sum(str_detect(D2_clean_base$artist, "\\\\")),
    sum(str_detect(D2_clean_base$artist, regex("\\sx\\s", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$artist, regex("\\s\\+\\s", ignore_case = TRUE)))
  )
) %>% 
  mutate(
    Percentage = round(Count / nrow(D2_clean_base) * 100, 1)
  )

d2_artist_stats %>% 
  gt() %>% 
  tab_header(title = "D2 Artist Field Pattern Statistics") %>% 
  fmt_number(columns = Count, decimals = 0, use_seps = TRUE)
```

### Representative Examples

```{r}
display_samples <- function(data, pattern_regex, label) {
  cat("\n=== Example:", label, "===\n")
  data %>% 
    filter(str_detect(artist, regex(pattern_regex, ignore_case = TRUE))) %>% 
    distinct(artist) %>% 
    slice_head(n = 2) %>% 
    pull(artist) %>% 
    print()
}

# Run the function for each pattern
display_samples(D2_clean_base, "Featuring", "Contains 'Featuring'")
display_samples(D2_clean_base, "feat\\.|ft\\.", "Contains 'feat.' or 'ft.'")
display_samples(D2_clean_base, "&", "Contains '&'")
display_samples(D2_clean_base, ",", "Contains comma ','")
display_samples(D2_clean_base, "[\\(\\)\\[\\]]", "Contains brackets () or []")
display_samples(D2_clean_base, "/", "Contains forward slash /")
display_samples(D2_clean_base, "\\\\", "Contains backslash \\")
display_samples(D2_clean_base, "\\sx\\s", "Contains 'x' or 'X'")
display_samples(D2_clean_base, "\\s\\+\\s", "Contains '+'")
```

### Cleaning Strategy

```{r}
cat("\n========================================\n")
cat("D2 Artist Field Cleaning Strategy\n")
cat("========================================\n\n")

cat("Core Principle: Conservative cleaning - Only remove explicit collaboration markers.\n\n")

cat("Data-Driven Rationale:\n")
cat("  - Featuring: 4.03%  -> Remove it and all following content.\n")
cat("  - feat./ft.: 0.02%  -> Remove it and all following content.\n")
cat("  - &:         4.34%  -> Preserve (part of the artist name).\n")
cat("  - Comma:     0.62%  -> Preserve (let featuring logic naturally handle it).\n")
cat("  - Brackets:  0.15%  -> Preserve (often artist aliases or metadata).\n")
cat("  - / + x:     <0.2%  -> Preserve (part of the artist name).\n\n")

cat("Cleaning Objectives:\n")
cat("  1. Remove 'Featuring' and all following content.\n")
cat("  2. Remove 'feat.' / 'ft.' and all following content.\n")
cat("  3. Preserve '&', '/', '+', 'x' (these are intrinsic to the artist names).\n\n")

cat("Exclusions (I Do Not Clean):\n")
cat("  - '&' symbol (may indicate a collaborative group name).\n")
cat("  - Brackets (may contain artist aliases, e.g., 'Ernie (Jim Henson)').\n")
cat("  - Commas (may be part of a band name, e.g., 'Blood, Sweat & Tears').\n")

```

------------------------------------------------------------------------

## 2.2 Track Column Pattern Analysis

```{r}
cat("\n========================================\n")
cat("D2 (Spotify) - Track Field Analysis\n")
cat("========================================\n\n")

d2_track_stats <- tibble(
  Pattern = c(
    "Contains Parentheses ()",
    "Contains Brackets []",
    "Contains ' - ' (dash)",
    "Contains 'feat.'",
    "Contains 'Remaster'",
    "Contains 'Live'",
    "Contains 'Remix'",
    "Contains 'Radio Edit'",
    "Contains 'feat' inside ()",
    "Contains 'Pt.' or 'Part'",
    "Contains 'Take'"
  ),
  Count = c(
    sum(str_detect(D2_clean_base$track, "\\(.*\\)")),
    sum(str_detect(D2_clean_base$track, "\\[.*\\]")),
    sum(str_detect(D2_clean_base$track, " - ")),
    sum(str_detect(D2_clean_base$track, regex("feat\\.", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$track, regex("Remaster", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$track, regex("Live", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$track, regex("Remix", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$track, regex("Radio Edit", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$track, regex("\\(.*feat", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$track, regex("Pt\\.|Part", ignore_case = TRUE))),
    sum(str_detect(D2_clean_base$track, regex("Take", ignore_case = TRUE)))
  )
) %>% 
  mutate(Percentage = round(Count / nrow(D2_clean_base) * 100, 1))

d2_track_stats %>% 
  gt() %>% 
  tab_header(title = "D2 Track Field Pattern Statistics") %>% 
  fmt_number(columns = Count, decimals = 0, use_seps = TRUE)
```

### Representative Examples

```{r}
# Reusing the helper function for clean output
show_track_examples <- function(data, pattern_regex, label) {
  cat("\n=== Example:", label, "===\n")
  data %>% 
    filter(str_detect(track, regex(pattern_regex, ignore_case = TRUE))) %>% 
    distinct(track) %>% 
    slice_head(n = 5) %>% 
    pull(track) %>% 
    print()
}

show_track_examples(D2_clean_base, "\\(.*\\)", "Contains Parentheses ()")
show_track_examples(D2_clean_base, "\\[.*\\]", "Contains Brackets []")
show_track_examples(D2_clean_base, " - ", "Contains ' - ' (dash)")
show_track_examples(D2_clean_base, "feat\\.", "Contains 'feat.'")
show_track_examples(D2_clean_base, "Remaster", "Contains 'Remaster'")
show_track_examples(D2_clean_base, "Live", "Contains 'Live'")
show_track_examples(D2_clean_base, "Remix", "Contains 'Remix'")
show_track_examples(D2_clean_base, "Radio Edit", "Contains 'Radio Edit'")
show_track_examples(D2_clean_base, "\\(.*feat", "Contains 'feat' inside ()")
show_track_examples(D2_clean_base, "Pt\\.|Part", "Contains 'Pt.' or 'Part'")
show_track_examples(D2_clean_base, "Take", "Contains 'Take'")
```

### Cleaning Strategy

```{r}
cat("\n========================================\n")
cat("D2 Track Field Cleaning Strategy\n")
cat("========================================\n\n")

cat("Core Principle: Active Dash Cleaning - D2 dash usage differs significantly from D1.\n\n")

cat("Data-Driven Rationale:\n")
cat("  - Parentheses ():   6.89%  -> Selective removal (target version tags only).\n")
cat("  - Brackets []:      0.22%  -> Remove all.\n")
cat("  - Dash -:           6.95%  -> Active cleaning (often indicates version tags).\n")
cat("  - Remaster:         2.10%  -> Remove (often found after dash).\n")
cat("  - Live:             1.47%  -> Remove (often found after dash).\n")
cat("  - Remix:            0.39%  -> Remove.\n")
cat("  - Radio Edit:       0.10%  -> Remove.\n")
cat("  - feat.:            0.36%  -> Remove (in parentheses or after dash).\n\n")

cat("Cleaning Objectives:\n")
cat("  1. Remove all brackets and their content.\n")
cat("  2. Remove version tags inside parentheses (remix, remaster, live, feat., etc.).\n")
cat("  3. Remove version tags following a dash (Remaster, Live, Remix, Radio Edit, etc.).\n\n")

cat("Exclusions (I Do Not Clean):\n")
cat("  - Standard parentheses (likely part of the title, e.g., 'Samba De Uma Nota So (One Note Samba)').\n")
cat("  - 'Part' or 'Pt.' following a dash (likely song identification, e.g., 'Hip City - Pt. 2').\n\n")

cat("Note on D2 vs. D1 Dash Differences:\n")
cat("  - D1 Dash: 98.5% are subtitles -> Conservative approach (remove Remix only).\n")
cat("  - D2 Dash: Frequent version tagging -> Active approach (remove Remaster, Live, Remix, etc.).\n")
```

------------------------------------------------------------------------

# Part 3: Dataset 3 (Music Dataset) Pattern Analysis

## 3.1 Artist Column Pattern Analysis

```{r}
cat("\n========================================\n")
cat("D3 (Music Dataset) - Column Field Analysis\n")
cat("========================================\n\n")

d3_artist_stats <- tibble(
  Pattern = c(
    "Contains 'Featuring'",
    "Contains 'feat.' or 'ft.'",
    "Contains '&'",
    "Contains comma ','",
    "Contains brackets () or []",
    "Contains forward slash /",
    "Contains backslash \\",
    "Contains 'x' or 'X' (Collab marker)",
    "Contains '+' (Collab marker)"
  ),
  Count = c(
    sum(str_detect(D3_clean_base$artist_name, regex("Featuring", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$artist_name, regex("feat\\.|ft\\.", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$artist_name, "&")),
    sum(str_detect(D3_clean_base$artist_name, ",")),
    sum(str_detect(D3_clean_base$artist_name, "[\\(\\)\\[\\]]")),
    sum(str_detect(D3_clean_base$artist_name, "/")),
    sum(str_detect(D3_clean_base$artist_name, "\\\\")),
    sum(str_detect(D3_clean_base$artist_name, regex("\\sx\\s", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$artist_name, regex("\\s\\+\\s", ignore_case = TRUE)))
  )
) %>% 
  mutate(
    Percentage = round(Count / nrow(D3_clean_base) * 100, 2)
  )

d3_artist_stats %>% 
  gt() %>% 
  tab_header(
    title = "D3 Artist Field Pattern Statistics"
  ) %>% 
  fmt_number(
    columns = Count,
    decimals = 0,
    use_seps = TRUE
  )
```

### Representative Examples

```{r}
# Reusing the display_samples helper function
# Make sure to use 'artist_name' column for D3
display_d3_samples <- function(data, pattern_regex, label) {
  cat("\n=== Example:", label, "===\n")
  data %>% 
    filter(str_detect(artist_name, regex(pattern_regex, ignore_case = TRUE))) %>% 
    distinct(artist_name) %>% 
    slice_head(n = 2) %>% 
    pull(artist_name) %>% 
    print()
}

display_d3_samples(D3_clean_base, "Featuring", "Contains 'Featuring'")
display_d3_samples(D3_clean_base, "feat\\.|ft\\.", "Contains 'feat.' or 'ft.'")
display_d3_samples(D3_clean_base, "&", "Contains '&'")
display_d3_samples(D3_clean_base, ",", "Contains comma ','")
display_d3_samples(D3_clean_base, "[\\(\\)\\[\\]]", "Contains brackets () or []")
display_d3_samples(D3_clean_base, "/", "Contains forward slash /")
display_d3_samples(D3_clean_base, "\\\\", "Contains backslash \\")
display_d3_samples(D3_clean_base, "\\sx\\s", "Contains 'x' or 'X'")
display_d3_samples(D3_clean_base, "\\s\\+\\s", "Contains '+'")
```

### Cleaning Strategy

```{r}
cat("\n========================================\n")
cat("D3 Artist Field Cleaning Strategy\n")
cat("========================================\n\n")

cat("Core Principle: Almost no cleaning required.\n\n")

cat("Data-Driven Rationale:\n")
cat("  - Featuring: 0.00%  -> Does not exist.\n")
cat("  - feat./ft.: 0.01%  -> These are part of the artist names, not collaboration markers.\n")
cat("  - &:         2.97%  -> Preserve (part of the artist name).\n")
cat("  - Comma:     0.79%  -> Preserve (part of band names).\n")
cat("  - Brackets:  0.01%  -> Preserve (part of band names).\n")
cat("  - / + x:     <0.2%  -> Preserve (part of artist names).\n\n")

cat("Cleaning Objectives:\n")
cat("  1. Only execute str_trim() to remove leading/trailing whitespace.\n")
cat("  2. Do not remove any content.\n\n")
```

------------------------------------------------------------------------

3.2 Track Column Pattern Analysis

```{r}
cat("\n========================================\n")
cat("D3 (Music Dataset) - Track Coulmn Analysis\n")
cat("========================================\n\n")

d3_track_stats <- tibble(
  Pattern = c(
    "Contains Parentheses ()",
    "Contains Brackets []",
    "Contains ' - ' (dash)",
    "Contains 'feat.'",
    "Contains 'Remaster'",
    "Contains 'Live'",
    "Contains 'Remix'",
    "Contains 'Radio Edit'",
    "Contains 'feat' inside ()",
    "Contains '+' (Part of title)",
    "Contains 'Pt.' or 'Part'",
    "Contains 'Take'"
  ),
  Count = c(
    sum(str_detect(D3_clean_base$track_name, "\\(.*\\)")),
    sum(str_detect(D3_clean_base$track_name, "\\[.*\\]")),
    sum(str_detect(D3_clean_base$track_name, " - ")),
    sum(str_detect(D3_clean_base$track_name, regex("feat\\.", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$track_name, regex("Remaster", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$track_name, regex("Live", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$track_name, regex("Remix", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$track_name, regex("Radio Edit", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$track_name, regex("\\(.*feat", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$track_name, regex("\\s\\+\\s", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$track_name, regex("Pt\\.|Part", ignore_case = TRUE))),
    sum(str_detect(D3_clean_base$track_name, regex("Take", ignore_case = TRUE)))
  )
) %>% 
  mutate(Percentage = round(Count / nrow(D3_clean_base) * 100, 2))

d3_track_stats %>% 
  gt() %>% 
  tab_header(title = "D3 Track Field Pattern Statistics") %>% 
  fmt_number(columns = Count, decimals = 0, use_seps = TRUE)
```

### Representative Examples

```{r}
show_d3_track_examples <- function(data, pattern_regex, label) {
  cat("\n=== Example:", label, "===\n")
  data %>% 
    filter(str_detect(track_name, regex(pattern_regex, ignore_case = TRUE))) %>% 
    distinct(track_name) %>% 
    slice_head(n = 5) %>% 
    pull(track_name) %>% 
    print()
}

show_d3_track_examples(D3_clean_base, "\\(.*\\)", "Contains Parentheses ()")
show_d3_track_examples(D3_clean_base, "\\[.*\\]", "Contains Brackets []")
show_d3_track_examples(D3_clean_base, " - ", "Contains ' - ' (dash)")
show_d3_track_examples(D3_clean_base, "feat\\.", "Contains 'feat.'")
show_d3_track_examples(D3_clean_base, "Remaster", "Contains 'Remaster'")
show_d3_track_examples(D3_clean_base, "Live", "Contains 'Live'")
show_d3_track_examples(D3_clean_base, "Remix", "Contains 'Remix'")
show_d3_track_examples(D3_clean_base, "Radio Edit", "Contains 'Radio Edit'")
show_d3_track_examples(D3_clean_base, "\\(.*feat", "Contains 'feat' inside ()")
show_d3_track_examples(D3_clean_base, "\\s\\+\\s", "Contains '+' (Part of title)")
show_d3_track_examples(D3_clean_base, "Pt\\.|Part", "Contains 'Pt.' or 'Part'")
show_d3_track_examples(D3_clean_base, "Take", "Contains 'Take'")
```

### Cleaning Strategy

```{r}
cat("\n========================================\n")
cat("D3 Track Field Cleaning Strategy\n")
cat("========================================\n\n")

cat("Core Principle: Only clean version tags inside parentheses; no dash handling required.\n\n")

cat("Data-Driven Rationale:\n")
cat("  - Parentheses ():   5.22%  -> Selective removal (target version tags only).\n")
cat("  - Brackets []:      0.08%  -> Remove all.\n")
cat("  - Dash -:           0.00%  -> Non-existent; no action needed.\n")
cat("  - feat.:            1.26%  -> Remove (almost entirely within parentheses).\n")
cat("  - Remaster:         0.08%  -> Remove (all within parentheses).\n")
cat("  - Live:             0.82%  -> Caution required (large number are lyric verbs).\n")
cat("  - Remix:            0.10%  -> Remove.\n")
cat("  - Radio Edit:       0.02%  -> Remove.\n\n")

cat("Cleaning Objectives:\n")
cat("  1. Remove all brackets and their content.\n")
cat("  2. Remove version tags inside parentheses (remix, remaster, feat., etc.).\n")
cat("  3. For 'Live', only remove explicit (live) formats inside parentheses.\n\n")

cat("Exclusions (I Do Not Clean):\n")
cat("  - Standard parentheses (likely part of the title).\n")
cat("  - 'live' used in lyrics ('as long as i live', 'i'm alive').\n")
cat("  - '+' symbol ('good + bad times' is the title itself).\n\n")
```
