---
title: "03 Data Wrangling"
author: "Fu Wei Hsu"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: right
    toc-title: "On this page"
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
---

# Executive Summary

**Purpose:** Standardize three datasets(data1,2,3) to enable join matching in subsequent steps.

**Tasks performed in this document:** 

- Artist field standardization (remove collaboration markers, unify case) 

- Song/Track field standardization (remove version markers, unify case) 

- Column name unification (align three datasets) - Generate Join Key (preparation for 04_Data_Joining)


------------------------------------------------------------------------

# Setup

```{r}
library(tidyverse)
library(janitor)
library(gt)

# Load cleaned data
D1_clean_base <- readRDS("../Data/cleaned_D1.rds")
D2_clean_base <- readRDS("../Data/cleaned_D2.rds")
D3_clean_base <- readRDS("../Data/cleaned_D3.rds")

cat("D1 (Billboard):", nrow(D1_clean_base), "rows\n")
cat("D2 (Spotify):", nrow(D2_clean_base), "rows\n")
cat("D3 (Music Dataset):", nrow(D3_clean_base), "rows\n")
```

------------------------------------------------------------------------

# Part 1: Dataset 1 (Billboard)

## Dataset 1 Wrangling

```{r}
# Artist cleaning function 
clean_d1_artist <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    # Remove parentheses containing featuring / feat. / duet with
    str_remove_all(regex("\\s*\\([^)]*(featuring|feat\\.|duet with)[^)]*\\)", ignore_case = TRUE)) %>%
    # Remove "featuring" and everything after it
    str_remove(regex("\\s+featuring.*$", ignore_case = TRUE)) %>%
    # Remove "feat" and everything after it
    str_remove(regex("\\s*\\(?\\s*feat\\..*$", ignore_case = TRUE)) %>%
    # Remove "ft." and everything after it
    str_remove(regex("\\s+ft\\..*$", ignore_case = TRUE)) %>%
    # Remove "with" and everything after it
    str_remove(regex("\\s+with\\s+.*$", ignore_case = TRUE)) %>%
    str_trim()
}

# Song cleaning function 
clean_d1_song <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    # Remove all square brackets and their contents
    str_remove_all("\\s*\\[.*?\\]") %>%
    # Remove version markers inside parentheses
    str_remove_all(regex("\\s*\\([^)]*remix[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*remaster[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*(live version|acoustic|unplugged)[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*(featuring|feat\\.|ft\\.)[^)]*\\)", ignore_case = TRUE)) %>%
    # Remove Remix marker after dash
    str_remove(regex("\\s*-\\s*.*remix.*$", ignore_case = TRUE)) %>%
    str_trim()
}

# Apply cleaning 
D1_wrangled <- D1_clean_base %>%
  mutate(
    artist_clean = clean_d1_artist(artist),
    song_clean   = clean_d1_song(song)
  )

cat("D1 cleaning data total rows:", nrow(D1_wrangled), "\n")
```

## Dataset 1 Validation

```{r}
# Artist changed records
D1_wrangled %>%
  filter(artist != artist_clean) %>%
  distinct(artist, artist_clean) %>%
  slice_head(n = 20) %>%
  gt() %>%
  tab_header(title = "D1 Artist Cleaning Effect (Changed Records Top 20)")
```

```{r}
# Artist changed records
D1_wrangled %>%
  filter(song != song_clean) %>%
  distinct(song, song_clean) %>%
  slice_head(n = 20) %>%
  gt() %>%
  tab_header(title = "D1 Song Cleaning Effect (Changed Records Top 20)")
```

```{r}
artist_changed <- D1_wrangled %>%
  filter(artist != artist_clean) %>%
  nrow()

song_changed <- D1_wrangled %>%
  filter(song != song_clean) %>%
  nrow()

cat("Artist records changed:", artist_changed, "rows (",
    round(artist_changed / nrow(D1_wrangled) * 100, 2), "%)\n")
cat("Song records changed:  ", song_changed, "rows (",
    round(song_changed / nrow(D1_wrangled) * 100, 2), "%)\n\n")

```

# Part 2: Dataset 2 (Spotify)

## Dataset 2 Wrangling

```{r d2-wrangling}
# Artist cleaning function
clean_d2_artist <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    # Remove parentheses containing featuring / feat. / duet with
    str_remove_all(regex("\\s*\\([^)]*(featuring|feat\\.|duet with)[^)]*\\)", ignore_case = TRUE)) %>%
    # Remove "featuring" and everything after it
    str_remove(regex("\\s+featuring.*$", ignore_case = TRUE)) %>%
    # Remove "feat." and everything after it
    str_remove(regex("\\s*\\(?\\s*feat\\..*$", ignore_case = TRUE)) %>%
    # Remove "ft." and everything after it
    str_remove(regex("\\s+ft\\..*$", ignore_case = TRUE)) %>%
    # Remove "with" and everything after it
    str_remove(regex("\\s+with\\s+.*$", ignore_case = TRUE)) %>%
    str_trim()
}

# Track cleaning function
clean_d2_track <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    # Remove version markers inside parentheses
    str_remove_all("\\s*\\[.*?\\]") %>%
    # Remove version markers inside parentheses
    str_remove_all(regex("\\s*\\([^)]*remix[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*remaster[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*(live version|acoustic|unplugged)[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*(featuring|feat\\.|ft\\.)[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*radio edit[^)]*\\)", ignore_case = TRUE)) %>%
    # Version markers after dash
    str_remove(regex("\\s*-\\s*.*(remaster|remastered|remasteris[eé]).*$", ignore_case = TRUE)) %>%
    str_remove(regex("\\s*-\\s*(live|version live|live version)\\s*$", ignore_case = TRUE)) %>%
    str_remove(regex("\\s*-\\s*.*remix.*$", ignore_case = TRUE)) %>%
    str_remove(regex("\\s*-\\s*radio edit.*$", ignore_case = TRUE)) %>%
    str_remove(regex("\\s*-\\s*feat\\..*$", ignore_case = TRUE)) %>%
    str_trim()
}

# Apply cleaning + unify column names
D2_wrangled <- D2_clean_base %>%
  rename(
    artist = artist,   
    song   = track     # rename track to song, aling with D1
  ) %>%
  mutate(
    artist_clean = clean_d2_artist(artist),
    song_clean   = clean_d2_track(song)
  )

cat("D2 cleaning data total rows:", nrow(D2_wrangled), "\n")
```

## Dataset 2 Validation

```{r}
#D2 Artist Cleaning Validation\n
D2_wrangled %>%
  filter(artist != artist_clean) %>%
  distinct(artist, artist_clean) %>%
  slice_head(n = 20) %>%
  gt() %>%
  tab_header(title = "D2 Artist Cleaning Effect (Changed Records Top 20)")
```

```{r}
#D2 Song Cleaning Validation
D2_wrangled %>%
  filter(song != song_clean) %>%
  distinct(song, song_clean) %>%
  slice_head(n = 20) %>%
  gt() %>%
  tab_header(title = "D2 Song Cleaning Effect (Changed Records Top 20)")
```

```{r}
# D2 Cleaning Summary
artist_changed <- D2_wrangled %>%
  filter(artist != artist_clean) %>%
  nrow()

song_changed <- D2_wrangled %>%
  filter(song != song_clean) %>%
  nrow()

cat("Artist records changed:", artist_changed, "rows (",
    round(artist_changed / nrow(D2_wrangled) * 100, 2), "%)\n")
cat("Songs records changed:", song_changed, "rows (",
    round(song_changed / nrow(D2_wrangled) * 100, 2), "%)\n\n")

```

# Part 3: Dataset 3 (Music Dataset)

## Dataset 3 Wrangling


```{r}
# D3 (Music Dataset) Wrangling
# Artist cleaning function 
clean_d3_artist <- function(x) {
  x %>%
    str_trim()
}

# Song cleaning function
clean_d3_song <- function(x) {
  x %>%
    str_trim() %>%
    # Remove all square brackets and their contents
    str_remove_all("\\s*\\[.*?\\]") %>%
    # Remove version markers inside parentheses
    str_remove_all(regex("\\s*\\([^)]*remix[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*remaster[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*(feat\\.|featuring)[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*radio edit[^)]*\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\([^)]*(acoustic|unplugged)[^)]*\\)", ignore_case = TRUE)) %>%
    # For "live", only remove explicit (live) or (live version) patterns in parentheses
    # Do not touch "live" appearing in song titles (e.g., 'as long as i live')
    str_remove_all(regex("\\s*\\(live version\\)", ignore_case = TRUE)) %>%
    str_remove_all(regex("\\s*\\(live\\)", ignore_case = TRUE)) %>%
    str_trim()
}

# Apply cleaning + unify column names
D3_wrangled <- D3_clean_base %>%
  rename(
    artist = artist_name,  # rename artist_name to artist, align with D1
    song   = track_name    # rename track_name to song, align with D1
  ) %>%
  mutate(
    artist_clean = clean_d3_artist(artist),
    song_clean   = clean_d3_song(song)
  )

cat("D3 cleaning data total rows:", nrow(D3_wrangled), "\n")
```

## Dataset 3 Validation

```{r}
# D3 artist
artist_changed <- D3_wrangled %>%
  filter(artist != artist_clean) %>%
  nrow()

cat("Artist records changed:", artist_changed, "rows\n")

D3_wrangled %>%
  filter(artist != artist_clean) %>%
  distinct(artist, artist_clean) %>%
  slice_head(n = 20) %>%
  gt() %>%
  tab_header(title = "3 Artist Cleaning Effect (Changed Records)")
```

```{r}
# D3 Song Cleaning Validation
D3_wrangled %>%
  filter(song != song_clean) %>%
  distinct(song, song_clean) %>%
  slice_head(n = 20) %>%
  gt() %>%
  tab_header(title = "D3 Song Cleaning Effect (Changed Records Top 20)")
```

```{r}
# D3 Cleaning Summary
artist_changed <- D3_wrangled %>%
  filter(artist != artist_clean) %>%
  nrow()

song_changed <- D3_wrangled %>%
  filter(song != song_clean) %>%
  nrow()

cat("Artist records changed: ", artist_changed, "rows (",
    round(artist_changed / nrow(D3_wrangled) * 100, 2), "%)\n")
cat("Song records changed: ", song_changed, "rows (",
    round(song_changed / nrow(D3_wrangled) * 100, 2), "%)\n\n")

```

# Part 4: Join Key Creation

## All Data create Join Key

```{r}
# Generate Join Key
D1_wrangled <- D1_wrangled %>%
  mutate(
    join_key = paste(artist_clean, song_clean, sep = "|")
  )

D2_wrangled <- D2_wrangled %>%
  mutate(
    join_key = paste(artist_clean, song_clean, sep = "|")
  )

D3_wrangled <- D3_wrangled %>%
  mutate(
    join_key = paste(artist_clean, song_clean, sep = "|")
  )

```

## D1 Join Key

```{r}
D1_wrangled %>%
  select(artist, song, artist_clean, song_clean, join_key) %>%
  distinct(join_key, .keep_all = TRUE) %>%
  slice_head(n = 5) %>%
  gt() %>%
  tab_header(title = "D1 Join Key")
```

## D2 Join Key

```{r join-key-d2-d3}
D2_wrangled %>%
  select(artist, song, artist_clean, song_clean, join_key) %>%
  distinct(join_key, .keep_all = TRUE) %>%
  slice_head(n = 5) %>%
  gt() %>%
  tab_header(title = "D2 Join Key")
```

## D3 Join Key

```{r}
D3_wrangled %>%
  select(artist, song, artist_clean, song_clean, join_key) %>%
  distinct(join_key, .keep_all = TRUE) %>%
  slice_head(n = 5) %>%
  gt() %>%
  tab_header(title = "D3 Join Key")
```

# Part 5: Data join key overlap predication

```{r}
# Before the formal join, quickly verify key overlap across the three datasets
# D1 unique join keys
d1_keys <- D1_wrangled %>% distinct(join_key) %>% pull(join_key)
d2_keys <- D2_wrangled %>% distinct(join_key) %>% pull(join_key)
d3_keys <- D3_wrangled %>% distinct(join_key) %>% pull(join_key)

d1_d2_overlap <- sum(d1_keys %in% d2_keys)
d1_d3_overlap <- sum(d1_keys %in% d3_keys)

cat("D1 unique keys:", length(d1_keys), "\n")
cat("D2 unique keys:", length(d2_keys), "\n")
cat("D3 unique keys:", length(d3_keys), "\n\n")

cat("D1 ∩ D2 overlap:", d1_d2_overlap, "rows (",
    round(d1_d2_overlap / length(d1_keys) * 100, 2), "% of D1)\n")
cat("D1 ∩ D3 overlap:", d1_d3_overlap, "rows (",
    round(d1_d3_overlap / length(d1_keys) * 100, 2), "% of D1)\n\n")
```

# Part 6 saveRDS

```{r}
saveRDS(D1_wrangled, "../Data/wrangled_D1.rds")
saveRDS(D2_wrangled, "../Data/wrangled_D2.rds")
saveRDS(D3_wrangled, "../Data/wrangled_D3.rds")
```

**Note** Due to formatting inconsistencies between D3 and the Billboard dataset, the exact match rate between D1 and D3 is approximately 10.6%. As a result, D3 is used primarily as a supplementary source for genre classification and lyrical sentiment analysis, rather than as a fully integrated dataset. Findings derived from the D1–D3 joined subset should be interpreted with caution given the limited coverage.
