---
title: "04 Data Joining.qmd"
author: "Fu Wei Hsu"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: right
    toc-title: "On this page"
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
---

# Executive Summary

**Purpose:** Merge three standardized datasets (data1,2,3) to produce the final dataset for subsequent analysis.

**Tasks performed in this document:**

- D1 + D2 exact match (primary join)

- D1+D2 match evaluation (match rate, decade distribution)

- D1 + D3 exact match (auxiliary join) - Output final dataset for analysis

**Data Source:** `03_Data_Wrangling.qmd`three RDSs

------------------------------------------------------------------------

```{r}
# Setup
library(tidyverse)
library(janitor)
library(gt)

# Load cleaned data from 03
D1_wrangled <- readRDS("../Data/wrangled_D1.rds")
D2_wrangled <- readRDS("../Data/wrangled_D2.rds")
D3_wrangled <- readRDS("../Data/wrangled_D3.rds")

cat("D1 (Billboard):", nrow(D1_wrangled), "rows\n")
cat("D2 (Spotify):", nrow(D2_wrangled), "rows\n")
cat("D3 (Music Dataset):", nrow(D3_wrangled), "rows\n\n")
```

## Part 1: D1 + D2 Join (Primary Join)

```{r}
# Prepare D2 join columns
# Keep one row per join_key in D2 (deduplication)
# Retain all audio features and target
D2_for_join <- D2_wrangled %>%
  distinct(join_key, .keep_all = TRUE)

cat("Original:", nrow(D2_wrangled), "rows, duplicate join_keys removed\n")

# Execute Left Join
# Use left join to keep all D1 records
# Matched records will include D2 audio features
# Unmatched records will have NA for audio feature columns
D1_D2_joined <- D1_wrangled %>%
  left_join(
    D2_for_join %>% select(
      join_key,
      # audio features
      danceability, energy, key, loudness, mode,
      speechiness, acousticness, instrumentalness,
      liveness, valence, tempo, duration_ms,
      time_signature, chorus_hit, sections,
      # target label
      target,
      # decade
      decade
    ),
    by = "join_key"
  )

#D1 + D2 Join complete
cat("Total rows after join:", nrow(D1_D2_joined), "\n")
```

## Part 2: D1+D2 Match Rate Evaluation

```{r}
# Overall Match Rate
matched    <- D1_D2_joined %>% filter(!is.na(target)) %>% nrow()
unmatched  <- D1_D2_joined %>% filter(is.na(target)) %>% nrow()
total      <- nrow(D1_D2_joined)

cat("Total records:    ", total, "rows\n")
cat("Matched:    ", matched, "rows (",
    round(matched / total * 100, 2), "%)\n")
cat("unmatched :      ", unmatched, "rows (",
    round(unmatched / total * 100, 2), "%)\n\n")

# Match Rate by Decade
D1_D2_joined %>%
  mutate(year = as.integer(format(date, "%Y")),
         decade_d1 = paste0(floor(year / 10) * 10, "s")) %>%
  group_by(decade_d1) %>%
  summarise(
    total     = n(),
    matched   = sum(!is.na(target)),
    match_pct = round(matched / total * 100, 2)
  ) %>%
  arrange(decade_d1) %>%
  gt() %>%
  tab_header(title = "D1+D2 Match Rate by Dacades") %>%
  fmt_number(columns = c(total, matched), decimals = 0, use_seps = TRUE)
```

## Part 3: D1-D2 Unmatched Records Sample Check

```{r}
D1_D2_joined %>%
  filter(is.na(target)) %>%
  mutate(year = as.integer(format(date, "%Y"))) %>%
  distinct(artist_clean, song_clean, year) %>%
  arrange(year) %>%
  slice_head(n = 10) %>%
  gt() %>%
  tab_header(title = "D1+D2 Unmatched Records Sample")
```

## Part 4: D1 + D2 + D3 Join (Auxiliary Join)

```{r}

D3_for_join <- D3_wrangled %>%
  distinct(join_key, .keep_all = TRUE)

cat("D3 distinct rows:", nrow(D3_for_join), "\n")

D1_D2_D3_joined <- D1_D2_joined %>%
  left_join(
    D3_for_join %>% select(
      join_key,
      genre,
      dating, violence, `world/life`, `night/time`,
      romantic, communication, obscene, music,
      sadness, feelings, topic
    ),
    by = "join_key"
  )


# Validation 
total      <- nrow(D1_D2_D3_joined)
d2_matched <- sum(!is.na(D1_D2_D3_joined$target))
d3_matched <- sum(!is.na(D1_D2_D3_joined$genre))
both_matched <- sum(!is.na(D1_D2_D3_joined$target) & 
                    !is.na(D1_D2_D3_joined$genre))

cat("Total rows: ", total, "\n")
cat("D2 matched (has target):  ", d2_matched,
    "(", round(d2_matched / total * 100, 2), "%)\n")
cat("D3 matched (has genre):   ", d3_matched,
    "(", round(d3_matched / total * 100, 2), "%)\n")
cat("Both D2 + D3 matched:     ", both_matched,
    "(", round(both_matched / total * 100, 2), "%)\n\n")
```

## Part 5: Save Final Analysis Dataset

```{r}
# Save single merged file
saveRDS(D1_D2_D3_joined, "../Data/D1_D2_D3_joined.rds")

cat("Dataset size check:\n")
cat("Total rows:", nrow(D1_D2_D3_joined), "\n")
cat("Total columns:", ncol(D1_D2_D3_joined), "\n")
```
